---
title: "Gov 50 Final Project"
author: "Joel Crawford"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to Congress. I want to look at relationship between previous years of Congress and the numbers of legislation introduced and passed in a particular congress. Is more legislation passed today then many years ago? I hypothesize that less legislation has been passed over time. My sample is compromised of each of the 93rd - 110th Congress. My unit of analysis is a Congress. The explanatory variables of interest is coded is the number of terms served counting the current term. If the variable coded = 7, it would mean that the member had served 7 terms in Congress. My outcome variable is the count of pieces of legislation introduced and pieces that became law. If the variable coded were = 4, this would mean that 4 pieces of legislation became law. The variables measured are easily accessible online via Congress.gov. I got the data set from the Center for Effective Lawmaking. If more senior members of congress introduced less pieces, but passed more legislation, then my hypothesis would be correct. However, if the data is run, and it appears that the more legislation becomes law in the present that in previous sessions of Congress, then my hypothesis would be very wrong.  



```{r}
library(readr)
library(gapminder)
library(tidyverse)
library(dplyr)
library(ggplot2)


CONGRESS_DATA_ <- read_csv("CONGRESS DATA .csv", 
    col_types = cols(congress = col_number(), 
        all_law = col_number()))

ggplot(CONGRESS_DATA_) + geom_col(aes(x = congress, y = all_law)) +
  labs(title = "The Number of Bills that become Law per Congress"
, x = "Session of Congress", 
       y = "Number of Bills that became Law")

```


```{r}

model<- lm(data = CONGRESS_DATA_, all_law~congress)|>
  broom::tidy()|>
  select(term, est=estimate, se= std.error, pval=p.value)

knitr::kable(model, digits=3, col.names = c("Variable", "Coefficeient Estimate", "Standard Error", "P-Value"))





```

From the plot entitled "The Number of Bills that became law", we can see that as the sessions Congress increase, the number of bills that become law decrease. In other words, in more recent years, there have been less Bills that have become law. From the Table, we can see that this relationship is significant, and that as the sessions of Congress increase, there is about a 2% decrase in bills that become law. 