---
title: "Gov 50 Final Project"
author: "Joel Crawford"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to Congress and the efficacy of legislation in preventing crime. Do Republican lead Congress have less crime? Or do Democratic lead Congress' lead to less crime. I will predict that when Democrats are in the majority, there is less crime than when Republicans are in the majority. My sample is compromised of each of the 93rd - 115h Congress. My unit of analysis is a Congress. The explanatory variables of interest is the political party in charge. If the variable coded = 0, it would mean that the Republicans are in the majority party in Congress. If the variable coded = 1, it would means that Democrats are in charge. My outcome variable is the number of violent offenses during each particular Congress. If the variable coded were = 4000, this would mean that, there were 4000 violent offenses for every 100,000 people in the population. 


To create this data, I had to merge data from two different places and essential create my own data set. I got the data regarding the party from the Brooking Institute, and I received the data regarding the crime rate in each session from the Federak Bureau of Investigation (FBI) Crime Data Explorer. 

If I run my data, and it shows that when Republicans are in the majority, there is less crime, then my hypothesis would be wrong. But, if the data shows the there is more crime when Republicans are in office, then my hypothesis would be right. 

```{r}
library(readr)
library(gapminder)
library(tidyverse)
library(dplyr)
library(readxl)
library(ggplot2)
```



```{r}

library(readxl)
congresssional_data_final <- read_excel("~/Downloads/congresssional data final.xlsx", 
    col_types = c("numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "text", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric"))



ggplot(congresssional_data_final, aes(x = party, y = crime, fill = party)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Comparison of Reported Crimes by Party in Charge",
    x = "Party in Charge",
    y = "Number Violent Crime Offenses
    per 100,000 people"
  ) +
  scale_fill_manual(values = c(R = "steelblue1", D = "indianred1"))

```

Looking at this data, we see a graph that shows the amount of crime over the entire time period (93rd Congress to 115th Congress). It breaks the crime down based on the party that is in charge. On the left, we see the over the entire time period, the Republican majority saw way more violent crimes, than the Democratic majority. 

```{r}
cor(congresssional_data_final$party_num, congresssional_data_final$crime)
```

This calculation confirms a positive linear relationship between the party and the amount crime. In other words, when Democrats are in the majority (party_num = 1), there is more crime. But, when Republicans are in office (party_num = 0), there is less crime. 



```{r}
model<- lm(data = congresssional_data_final, crime~party_num)|>
  broom::tidy()|>
  select(term, est=estimate, se= std.error, pval=p.value)

knitr::kable(model, digits=3, col.names = c("Variable", "Coefficient Estimate", "Standard Error", "P-Value"))
```

So, this graph shows us a few different things. First of all, the intercept is when party_num = 0, or when Republicans are in power. This graph shows us that when Republicans are in the majority, the estimated amount of crime is 1,475,862.2. The p-value is 0, meaning that this is statically significant.

In addition, the graph tells us that when party_num = 1, or when Democrats are in power, there is estimated to be an increase of 210,772.6 cases of crime, compared to when Republicans are in power. The p-value of 0.01 tells us that this is statistically significant. 

Overall, this graph shows us that when democrats are in power, it is estimated that there will be substantially more cases of crime. 



Conclusion


In conclusion, my hypothesis was incorrect. When democrats are in power, the data shows us that there tends to be more crime, as opposed to when Republicans are in power. In this data set, there are various limitations that I would fix if I had more data or more money. Mainly, I would like to look at the types of bills that are being passed by each party to see if they are directly working to decrease the amount of crime. In other words, are there specific bills that are being passed that can explain why a Republican majority has less crime. In addition, since it takes time for policies to set in, it might also be that the decreased crime is due to Democrat polices. We would need more data to see if the Republicans are intentionally doing things to decrease the crime, or if the crime is decreasing due to policies implemented prior to them taking the majority.  


```{r}
cor(congresssional_data_final$party_num, congresssional_data_final$crime)
```





```{r}
ggplot(congresssional_data_final, aes(x = crime_rate, y = congress_in, color = party)) +
  geom_point() +
  labs(
    title = "Scatter Plot of Crime and Party in Charge",
    x = "Crime Reported",
    y = "Congress Session"
  ) +
  scale_color_manual(values = c(R = "steelblue1", D = "indianred1"))

```

```{r}
cor(congresssional_data_final$crime_rate, congresssional_data_final$congress_in)
```



```{r}

ggplot(congresssional_data_final, aes(x = congress_in, y = crime, color = party)) +
  geom_point() +
  labs(
    title = "Relationship between Crime and Party in Charge in each Congress",
    x = "Congress",
    y = "Number of Crimes Reported"
  ) +
  scale_color_manual(values = c(R = "steelblue1", D = "indianred1"))

```
```{r}
cor(congresssional_data_final$party_num, congresssional_data_final$crime)
```

But, if we look further at this data, we see that it is much more than just an increase in legislative activity, but additionally party in charge, that has an effect on the crime. It appears that a Republican Majority





```{r}

ggplot(congresssional_data_final, aes(x = party, y = crime, fill = party)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Crime by Majority Party in Congress",
    x = "Majority Party",
    y = "Number of Crimes"
  ) +
  scale_fill_manual(values = c(D = "steelblue1", R = "indianred1"))

```



```{r}
model<- lm(data = congresssional_data_final, crime~party_num)|>
  broom::tidy()|>
  select(term, est=estimate, se= std.error, pval=p.value)

knitr::kable(model, digits=3, col.names = c("Variable", "Coefficeient Estimate", "Standard Error", "P-Value"))
```



